```{r}
library(dplyr)
```

# Quiz 1

## Question 1

Consider the `mtcars` data set. Fit a model with mpg as the outcome that
includes number of cylinders as a factor variable and weight as confounder.
Give the adjusted estimate for the expected change in mpg comparing 8 cylinders
to 4.

### Answer:
```{r}
grouped.cars <- mtcars %>% group_by(as.factor(cyl))
```

```{r}
fit.result <- with(grouped.cars, lm(mpg ~ wt+as.factor(cyl)))
fit.result
fit.result$coefficients['as.factor(cyl)8']
```


## Question 2

Consider the mtcars data set. Fit a model with mpg as the outcome that includes
number of cylinders as a factor variable and weight as a possible confounding
variable.
Compare the effect of 8 versus 4 cylinders on mpg for the adjusted and
unadjusted by weight models. Here, adjusted means including the weight variable
as a term in the regression model and unadjusted means the model without weight
included.
What can be said about the effect comparing 8 and 4 cylinders after looking at
models with and without weight included?.

### Answer:

First, look at the regression where the weight is considered as a confounding
factor. This is the same as question 1.
```{r}
fit.result <- with(grouped.cars, lm(mpg ~ wt+as.factor(cyl)))
fit.result
fit.result$coefficients['as.factor(cyl)8']
```

Now, compare this with the scenario where weight isn't considered.
```{r}
fit.result <- with(grouped.cars, lm(mpg ~ as.factor(cyl)))
fit.result
fit.result$coefficients['as.factor(cyl)8']
```

When the weight is considered as a confounding factor, the effect of the
number of cylinders appears to decrease.

## Question 3

Consider the mtcars data set. Fit a model with mpg as the outcome that considers
number of cylinders as a factor variable and weight as confounder.
Now fit a second model with mpg as the outcome model that considers the
interaction between number of cylinders (as a factor variable) and weight.
Give the P-value for the likelihood ratio test comparing the two models and
suggest a model using 0.05 as a type I error rate significance benchmark.

### Answer:

First, look at the regression where the weight is considered as a confounding
factor. This is the same as question 1.
```{r}
fit.result.no.interaction <- with(grouped.cars, lm(mpg ~ wt+as.factor(cyl)))
summary(fit.result.no.interaction)
```

```{r}
fit.result.with.interaction <- with(grouped.cars, lm(mpg ~ wt+as.factor(cyl)+wt*as.factor(cyl)))
summary(fit.result.with.interaction)
```

```{r}
likelihood.ratio.test <- anova(fit.result.no.interaction,
                               fit.result.with.interaction)
likelihood.ratio.test
```

The interaction term between the 8 cylinder subset and the weight is
`r likelihood.ratio.test['Pr(>F)'][2,]`.
The difference is not significant, so we fail to reject the model with no
interaction term.

## Question 4

Consider the `mtcars` data set. Fit a model with mpg as the outcome that
includes number of cylinders as a factor variable and weight inlcuded in the
model as
```
lm(mpg ~ I(wt * 0.5) + factor(cyl), data = mtcars)
```
How is the wt coefficient interpretted?


### Answer:

```{r}
lm(mpg ~ I(wt * 0.5) + factor(cyl), data = mtcars)
```

Since the `wt` column is given in units of 1000 lbs, the above linear model
gives "The estimated expected change in MPG per one ton increase in weight for
a specific number of cylinders (4, 6, 8)."

I'm not sure I completely buy this because the same slope is used for all number
of cylinders.


## Question 5

Consider the following data set

```{r}
x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)
```

Give the hat diagonal for the most influential point

### Answer:

```{r}
fit <- lm(y~x)
plot(y~x)
abline(fit)
```

```{r}
hatvalues(fit)
```

## Question 6

Consider the following data set

```{r}
x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)
```

Give the slope dfbeta for the point with the highest hat value.

### Answer:

```{r}
dfbetas(fit)
```

## Question 7

Consider a regression relationship between Y and X with and without adjustment
for a third variable Z. Which of the following is true about comparing the
regression coefficient between Y and X with and without adjustment for Z.

### Answer:

For the the coefficient to change sign, there must be a significant interaction term.