# Quiz 4

## Question 1

Consider the space shuttle data `?shuttle` in the `MASS` library.
Consider modeling the use of the autolander as the outcome (variable name `use`).
Fit a logistic regression model with autolander (variable auto) use
(labeled as "auto" 1) versus not (0) as predicted by wind sign (variable wind).
Give the estimated odds ratio for autolander use comparing head winds, labeled
as "head" in the variable headwind (numerator) to tail winds (denominator).

### Answer:

```{r}
library(dplyr)
library(MASS)
my.shuttle <- shuttle %>% mutate(use = factor(use, levels(use)[c(2,1)]))
head(my.shuttle)
levels(my.shuttle$use)
levels(my.shuttle$wind)
```

```{r}
mdl <- glm(use ~ wind, 'binomial', data=my.shuttle)
summary(mdl)
```

```{r}
odds.head <- exp(mdl$coefficients[['(Intercept)']])
odds.tail <- exp(mdl$coefficients[['(Intercept)']] + mdl$coefficients[['windtail']])
odds.ratio <- odds.head/odds.tail
```
The odds ratio is `r round(odds.ratio, 5)`.

#### Cross check
```{r}
num.headwind <- nrow(shuttle[shuttle$wind == 'head',])
num.headwind.use <- nrow(shuttle[shuttle$wind == 'head' & shuttle$use == 'auto',])

num.tailwind <- nrow(shuttle[shuttle$wind == 'tail',])
num.tailwind.use <- nrow(shuttle[shuttle$wind == 'tail' & shuttle$use == 'auto',])

odds.head.check <- num.headwind.use/(num.headwind-num.headwind.use)
odds.tail.check <- num.tailwind.use/(num.tailwind-num.tailwind.use)
odds.head.check/odds.tail.check
```

## Question 2

Consider the previous problem.
Give the estimated odds ratio for autolander use comparing head winds
(numerator) to tail winds (denominator) adjusting for wind strength from the
variable magn.

### Answer:

```{r}
my.shuttle <- my.shuttle %>% mutate(magn = factor(magn, levels(magn)[c(1,2,4,3)]))
```

```{r}
mdl <- glm(use ~ wind + magn, 'binomial', data=my.shuttle)
summary(mdl)
```

```{r}
odds.head <- exp(mdl$coefficients[['(Intercept)']])
odds.tail <- exp(mdl$coefficients[['(Intercept)']] + mdl$coefficients[['windtail']])
odds.ratio <- odds.head/odds.tail
```
The odds ratio is `r round(odds.ratio, 5)`.

## Question 3

If you fit a logistic regression model to a binary variable, for example use of
the autolander, then fit a logistic regression model for one minus the outcome
(not using the autolander) what happens to the coefficients?

### Answer:

Flip sign.

#### Cross check
Recall the regression with auto = 2, and no auto = 1 gave.

```{r}
glm(use ~ wind, 'binomial', data=my.shuttle)$coef
```

```{r}
glm(use ~ wind, 'binomial', data=shuttle)$coef
```

## Question 4

Consider the insect spray data `InsectSprays`.
Fit a Poisson model using spray as a factor level. Report the estimated relative
rate comapring spray A (numerator) to spray B (denominator).

### Answer:

```{r}
mdl <- glm(count ~ spray, 'poisson',
           data=InsectSprays[InsectSprays$spray == 'A' |
                               InsectSprays$spray == 'B',])
mdl$coef
```

```{r}
rate.ratio <- exp(mdl$coefficients[1])/exp(mdl$coefficients[1] +
                                             mdl$coefficients[2])
```

The ratio of the rates is `r round(rate.ratio, 5)`.

## Question 5

Consider a Poisson glm with an offset, $t$. So, for example, a model of the form
`glm(count ~ x + offset(t), family = poisson)` where `x` is a factor variable
comparing a treatment (1) to a control (0) and `t` is the natural log of a
monitoring time.
What is impact of the coefficient for `x` if we fit the model
`glm(count ~ x + offset(t2), family = poisson)` where `t2 <- log(10) + t`?
In other words, what happens to the coefficients if we change the units of the
offset variable.
(Note, adding log(10) on the log scale is multiplying by 10 on the original
scale.)

### Answer:

The coefficients should remain unchanged.

## Question 6

Consider the data

```{r}
x <- -5:5
y <- c(5.12, 3.93, 2.67, 1.87, 0.52, 0.08, 0.93, 2.05, 2.54, 3.87, 4.97)
```

Using a knot point at 0, fit a linear model that looks like a hockey stick with
two lines meeting at x=0.
Include an intercept term, x and the knot point term. What is the estimated
slope of the line after 0?

### Answer:

```{r}
plot(x, y, pch = 21, bg='lightblue', cex=2)
```

```{r}
spline = (x>0)*x
rhs <- cbind(x, spline)
fit <- lm(y ~ rhs)
summary(fit)
yhat <- predict(fit)
plot(x, y, pch = 21, bg='lightblue', cex=2)
lines(x, yhat, col='red', lwd=2)
```

The slope after $x=0$ is
`r round(fit$coef['rhsx'] + fit$coef['rhsspline'], 5)`.
