# Quiz 4

## Question 1
A pharmaceutical company is interested in testing a potential blood pressure
lowering medication. Their first examination considers only subjects that
received the medication at baseline then two weeks later. The data are as
follows (SBP in mmHg)

| Subject  | Baseline	| Week 2 |
----------------------------
| 1 |	140	| 132 |
| 2	| 138	| 135 |
| 3	| 150	| 151 |
| 4	| 148	| 146 |
| 5	| 135	| 130 |

Consider testing the hypothesis that there was a mean reduction in blood
pressure? Give the P-value for the associated two sided T test.

Answer:
```{r}
blood.pressure <- data.frame(baseline=c(140, 138, 150, 148, 135),
                             week.2=c(132, 135, 151, 146, 130))
print(blood.pressure)
result <- t.test(blood.pressure$baseline, blood.pressure$week.2,
                 paired=TRUE,
                 alternative='two.sided')
print(result)
```


## Question 2
A sample of 9 men yielded a sample average brain volume of 1,100cc and a
standard deviation of 30cc. What is the complete set of values of μ0 that a
test of H0:μ=μ0 would fail to reject the null hypothesis in a two sided 5%
Students t-test?

Answer:
```{r}
sample.mean <- 1100
sample.sd <- 30
n.samples <- 9

sample.mean +c(-1,+1)*qt(0.975, n.samples-1)*sample.sd/sqrt(n.samples)
```

## Question 3
Researchers conducted a blind taste test of Coke versus Pepsi. Each of four
people was asked which of two blinded drinks given in random order that they
preferred. The data was such that 3 of the 4 people chose Coke. Assuming that
this sample is representative, report a P-value for a test of the hypothesis
that Coke is preferred to Pepsi using a one sided exact test.

Answer:
```{r}
n.samples <- 4
n.coke <- 3
pbinom(n.coke-1, n.samples, 0.5, lower.tail=FALSE)
```

Alternatively,

```{r}
binom.test(n.coke, n.samples, alternative='greater')
```


## Question 4
Infection rates at a hospital above 1 infection per 100 person days at risk are
believed to be too high and are used as a benchmark. A hospital that had
previously been above the benchmark recently had 10 infections over the last
1,787 person days at risk. About what is the one sided P-value for the relevant
test of whether the hospital is *below* the standard?

Answer:
```{r}
n.samples <- 1787
n.infections <- 10
benchmark <- 1/100

pbinom(n.infections, n.samples, benchmark)
```

## Question 5
Suppose that 18 obese subjects were randomized, 9 each, to a new diet pill and
a placebo. Subjects’ body mass indices (BMIs) were measured at a baseline and
again after having received the treatment or placebo for four weeks. The
average difference from follow-up to the baseline (followup - baseline) was
−3 kg/m2 for the treated group and 1 kg/m2 for the placebo group. The
corresponding standard deviations of the differences was 1.5 kg/m2 for the
treatment group and 1.8 kg/m2 for the placebo group. Does the change in BMI
appear to differ between the treated and placebo groups? Assuming normality
of the underlying data and a common population variance, give a pvalue for a
two sided t test.

Answer:
```{r}
n.sample.group <- 9
mean.diff.treated <- -3
mean.diff.placebo <- 1
sd.treated <- 1.5
sd.placebo <- 1.8

se.est <- sqrt(sd.treated^2/n.sample.group + sd.placebo^2/n.sample.group)
test.stat <- (mean.diff.treated - mean.diff.placebo)/se.est

pt(test.stat, 2*n.sample.group-1)
```


## Question 6
Brain volumes for 9 men yielded a 90% confidence interval of 1,077 cc to
1,123 cc. Would you reject in a two sided 5% hypothesis test of H0:μ=1,078?

Answer:
We would not reject the two sided 5% hypothesis test of H0: \mu=1078 since this
test is interested in cehcking the 95% confidence interval which fully includes
the 90% confidence interval


## Question 7
Researchers would like to conduct a study of 100 healthy adults to detect a
four year mean brain volume loss of .01 mm3. Assume that the standard deviation
of four year volume loss in this population is .04 mm3. About what would be the
power of the study for a 5% one sided test versus a null hypothesis of no
volume loss?

Answer:
```{r}
n.samples <- 100
alternate.mean <- 0.01
null.mean <- 0.
sd <- 0.04

error <- sd/sqrt(n.samples)
cut.value <- null.mean + qt(0.95, n.samples-1)*error

test.stat <- (alternate.mean - cut.value)/error

pt(test.stat, n.samples-1)
```

Or alternatively:
```{r}
power.t.test(n=n.samples, delta=(alternate.mean-null.mean), sd=sd,
             sig.level=0.05, alternative='one.sided', type='one.sample')
```


## Question 8
Researchers would like to conduct a study of n healthy adults to detect a four year mean brain volume loss of .01 mm3. Assume that the standard deviation of four year volume loss in this population is .04 mm3. About what would be the value of n needded for 90% power of type one error rate of 5% one sided test versus a null hypothesis of no volume loss?

Answer:
```{r}
power.t.test(power=0.90, delta=(alternate.mean-null.mean), sd=sd,
             sig.level=0.05, alternative='one.sided', type='one.sample')
```


## Question 9
As you increase the type one error rate, α, what happens to power?

Answer:
You will get larger power.