# Quiz 1

## Question 1
In this example, we want to know the probability of the mother having a disease
given the following known probabilities:
```{r}
p.either <- 0.17
p.both <- 0.06
p.father <- 0.12
```

We know the following

```
p.either = p.father + p.mother - p.both
```

We can rearange that equation to get 

```{r}
p.mother <- p.either + p.both - p.father
print(p.mother)
```

## Question 2
Suppose we have a uniform distribution between 0 and 1. What is the 75th
percentile?

```{r}
qunif(0.75, 0, 1)
```

## Question 3
You and a friend are playing a game where you flip a coin. Every time the coin
lands on heads, she gives you `X`. Every time the coin lands on tails, she gives
you `Y`. The probability of the coin landing on heads is `p`. What is the
relation between `X` and `Y` to mak the game fair (expected earnings = 0).
```
expected.earnings = p*X - (1-p)*Y = 0
```

Rearranging this, we get
```
p/(1-p) = Y/X
```

## Question 4
A distribution which is symmetric about zero must have a median of zero.

## Question 5
Consider the following PMF:
```{r}
x <- 1:4
p <- x/sum(x)
temp <- rbind(x, p)
rownames(temp) <- c("X", "Prob")
temp
```
What is its mean?
```{r}
sum(temp['X',]*temp['Prob',])
```

## Question 6
Suppose a pregnancy test has a sensitivity of 0.75 and specificity of 0.54.
Of the women in the study 30% are pregnant. Our subject received a positive
result on the test, what is her probability of being pregnant?

```{r}
sensitivity <- 0.75
specificity <- 0.52
prior <- 0.3
```

We can use Bayes theorem to solve this:
```
p(preg|+) = p(+|preg)*p(preg)/p(+)
p(preg|+) = p(+|preg)*p(preg)/(p(+|pregnant)*p(pregnant) + p(+|not pregnant)*p(not pregnant))
p(preg|+) = p(+|preg)*p(preg)/(p(+|pregnant)*p(pregnant) + (1-p(-|not pregnant))*(1-p(pregnant)))
p(preg|+) = sensitivity*prior/(sensitivity*prior + (1-specificity)*(1-prior))
```

```{r}
sensitivity*prior/(sensitivity*prior + (1-specificity)*(1-prior))
```