---
title: "Effect of various attributes on miles per gallon"
author: 'Brett Jackson'
# output: pdf_document
output: 
  pdf_document
---
```{r, echo=FALSE}                                                               
# Set the output options for large and small numbers                             
options(scipen = 5, digits = 2)
```                            

```{r, load_libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(GGally)
```

```{r, import.dataset, echo=FALSE}
cars <- mtcars %>%
        mutate(Trans=as.factor(am), cyl=as.factor(cyl), vs=as.factor(vs),
               gear=as.factor(gear)) %>%
        group_by(Trans) %>%
        select(-am, -qsec, -carb, -gear, -drat, -vs)
levels(cars$Trans)[1] = 'Auto'
levels(cars$Trans)[2] = 'Manual'
```

```{r, simple.summary, echo=FALSE}
simple.summary <- summarize(cars, mean(mpg))
```

```{r, simple.test, echo=FALSE}
simple.test <- t.test(cars[cars$Trans=='Manual',]$mpg,
                      cars[cars$Trans=='Auto',]$mpg,
                      alternative='greater')
```

```{r, define.basic.fit, echo=FALSE}
fit <- lm(mpg ~ as.factor(Trans), data=cars)
```

```{r, define.final.fit, echo=FALSE}
final.fit.result <- update(fit, mpg ~ as.factor(Trans) + wt + cyl + wt*as.factor(Trans))
```

## Executive summary
The goal of this study is to use the `mtcars` data set to investigate any
potential relationships associated with the fuel efficiency.
In particular, is there any significant difference in fuel efficiency
associated with the type of transmission (manual or automatic)?

By performing a linear regression analysis to build a descriptive model of the
`mtcars` dataset, accounting for the correlations between the car attributes,
it is shown that among the cars in the dataset, the cars with a manual
tranmsission do in fact have better fuel efficiency than those with an automatic
transmission, with a difference of
`r final.fit.result$coefficients[['as.factor(Trans)Manual']]` $\pm$
`r summary(final.fit.result)$coefficients['as.factor(Trans)Manual','Std. Error']`
miles per gallon.
The model includes the type of transmission, the car weight, and the number of
cylinders in the car engine. Other attributes are shown to provide little
additional descriptive power.

## Development of the model
The desriptive model is a linear regression of the various attributes of the
cars in the `mtcars` dataset.
Several variables are removed from the model as they have little correlation
with the fuel efficiency.

```{r, ref.label='import.dataset', eval=FALSE}
```

A pairs plot of the attributes in the data set is shown in the appendix.
The model is developed by, first including all the candidated attributes
and their interactions, and successively removing attributes from the model
whos coefficients are shown to be insignificant.
A cross check is perfirmed using a nested model test.

The initial model includes all the relevent attributes and their interaction
terms.

```{r, ref.label='define.basic.fit', eval=FALSE}
```

```{r}
summary(update(fit, mpg ~ as.factor(Trans) + wt + cyl + hp +
                 as.factor(Trans)*wt + as.factor(Trans)*cyl +
                 as.factor(Trans)*hp + wt*cyl + wt*hp + cyl*hp))$coefficients
```

It is clear that many terms have coefficients which are not significant
based on the $p$-value.
Terms were successively remvoved from the model, starting with least significant
coefficients, until the final model has all signifant terms.

```{r, ref.label='define.final.fit', eval=FALSE}
```

```{r}
summary(final.fit.result)$coefficients
```

The important attributes for this model are the transmission type, the weight,
and the number of cylinders, and the interaction between the transmission type
and the weight.
The resudiduals obtained by comparing the predictions from this model with
the observed fuel efficiencies in the `mtcars` dataset are shown here.

```{r, fig.height=3}
hist(residuals(final.fit.result), breaks = (-3:4)*2, col='steelblue',
     xlab='Resdiual', main='Model residuals')
```

## Interpretation
This model suggests that a difference of
`r final.fit.result$coefficients[['as.factor(Trans)Manual']]` $\pm$
`r summary(final.fit.result)$coefficients['as.factor(Trans)Manual','Std. Error']`
miles per gallon can be attributed to the type of transmission, where cars with an automatic
transmission tend to have lower fuel efficiency.

## Appendix

```{r, exploratory_plots, echo=TRUE, warning=FALSE, message=FALSE, fig.height=10, fig.width=10}
cars.sub <- cars[,c('mpg', 'wt', 'cyl', 'Trans')]
cars.sub <- cars
pairs.plot <- ggpairs(cars.sub,
                      color='Trans',
                      alpha=0.4,
                      upper = list(params = c(corrSize=1)))

corr.font.size <- 3
cars.labels <- labels(cars.sub)[[2]]
for (itx in seq_along(cars.labels)) {
  for (ity in seq_along(cars.labels)) {
    if (itx <= ity) {
      next
    }
    x.label <- cars.labels[itx]
    y.label <- cars.labels[ity]
    if (class(cars[[x.label]]) == 'factor' ||
        class(cars[[y.label]]) == 'factor') {
      next
    }
    tmp.aes <- aes_string(x=x.label, y=y.label, color='Trans')
    corr.plot <- ggally_cor(cars.sub,
                            tmp.aes,
                            size=corr.font.size)

    pairs.plot <- putPlot(pairs.plot, corr.plot, ity, itx)
  }
}

pairs.plot +
  theme(text = element_text(size=12),
        axis.text.x = element_text(angle=0, size=10),
        axis.text.y = element_text(angle=0, size=10))
```
